Follow-up Samples

### 2016-12-05 submission/manuscript_outline_20161024.Rmd
updated manuscript w/ newer abstract

source('code/functions.R')
source('code/graphFunctions.R')

loadLibs(c("pROC","randomForest","AUCRF", "Boruta", "dplyr", "tidyr", "ggplot2", "reshape2", 
           "gridExtra", "scales", "wesanderson", "vegan", "knitr"))

@@

```{r rdata_tables}
# Extra data and meta data needed
good_metaf <- read.csv('results/tables/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
metaI <- read.csv('results/tables/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
total_sub_Seqs <- read.delim('data/process/final.0.03.subsample.shared', header = T) %>% 
  select(contains("Otu0")) %>% rowSums()

# initial microbiome and FIT description data 
change_summary <- read.csv('results/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
alpha_summary <- read.csv('results/tables/alpha_table_summary.csv', header = T, row.names = 1)
beta_summary <- read.csv('results/tables/beta_diver_summary.csv', header = T, row.names = 1)
time_pvalues <- read.csv("results/tables/time_pvalues.csv", header = T, row.names = 1)

# training of model data
model_train_summary <- read.csv('results/tables/ROC_model_summary.csv', header = T)
model_test_AUC_summary <- read.csv('results/tables/auc_summary.csv', header = T, row.names = 1)
imp_model_vars <- read.csv('results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
rf_otu_tax <- read.csv('results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
rf_lowest_ID <- createTaxaLabeller(rf_otu_tax)

# model prediction and adjustment performance
IF_model_wilcox_paired_summary <- read.csv('results/tables/IF_model_wicox_paired_pvalue_summary.csv', header = T, row.names = 1)
confusion_summary <- read.csv('results/tables/confusion_summary.csv', header = T, row.names = 1)
follow_up_prob_summary <- read.csv('results/tables/follow_up_probability_summary.csv', header = T)
followup_cm <- read.csv('results/tables/followup_confusion_matrix.csv', header = T, row.names = 1)
initial_cm <- read.csv('results/tables/initial_confusion_matrix.csv', header = T, row.names = 1)
roc_pvalues <- read.csv('results/tables/roc_pvalue_summary.csv', header = T, row.names = 1)

# Predictive of both initial and follow up and in lesion model
pvalue_common_imp_vars <- read.csv('results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)

@@

Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.  Previous research suggests that the bacterial microbiome can be used as a biomarker for colorectal cancer (CRC).  These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.  In this study, we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for lesion (adenoma or carcinoma) after surgical treatment.  This model was tested on a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment. The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.  : For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```).  The top 5 most important operational taxonomic units (OTUs) for prediction of lesion classified to ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.  In the test set, the prediction of lesion for the initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).  Overall, our model accurately predicted those with an adenoma or carcinoma and following treatment it also decreased the positive probability of having a lesion.

@@

## Discussion
In this study we show that the overall community structure shows very little change between controls and those with either adenoma or carcinoma [Table S1] in our larger training set.  With respect to the follow up sample group there was very little difference in thetayc distance metric between those with adenoma or carcinoma[Figure 1a].  FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].  When investigating the bacterial community, specifically between these individuals, there was very little change in overall community in the adenoma group but there was a change in the carcinoma group [Figure 2].  This information showed that the removal and treatment of carcinoma for these cursory metrics had the largest response.  Yet these metrics alone would not be able to identify differences in adenomas.

>>>>

source('../code/functions.R')
source('../code/graphFunctions.R')

loadLibs(c("dplyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))

@@

```{r rdata_tables}
# Extra data and meta data needed
good_metaf <- read.csv('../results/tables/mod_metadata/good_metaf_final.csv', header = T, stringsAsFactors = F)
metaI <- read.csv('../results/tables/mod_metadata/metaI_final.csv', header = T, stringsAsFactors = F)
total_sub_Seqs <- read.delim('../data/process/final.0.03.subsample.shared', header = T) %>% 
  select(contains("Otu0")) %>% rowSums()

# initial microbiome and FIT description data 
change_summary <- read.csv('../results/tables/change_theta_fit_summary.csv', header = T, row.names = 1)
alpha_summary <- read.csv('../results/tables/alpha_table_summary.csv', header = T, row.names = 1)
beta_summary <- read.csv('../results/tables/beta_diver_summary.csv', header = T, row.names = 1)
time_pvalues <- read.csv("../results/tables/time_pvalues.csv", header = T, row.names = 1)

# training of model data
model_train_summary <- read.csv('../results/tables/ROC_model_summary.csv', header = T)
model_test_AUC_summary <- read.csv('../results/tables/auc_summary.csv', header = T, row.names = 1)
imp_model_vars <- read.csv('../results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
rf_otu_tax <- read.csv('../results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
rf_lowest_ID <- createTaxaLabeller(rf_otu_tax)

# model prediction and adjustment performance
IF_model_wilcox_paired_summary <- read.csv('../results/tables/IF_model_wicox_paired_pvalue_summary.csv', header = T, row.names = 1)
confusion_summary <- read.csv('../results/tables/confusion_summary.csv', header = T, row.names = 1)
follow_up_prob_summary <- read.csv('../results/tables/follow_up_probability_summary.csv', header = T)
followup_cm <- read.csv('../results/tables/followup_confusion_matrix.csv', header = T, row.names = 1)
initial_cm <- read.csv('../results/tables/initial_confusion_matrix.csv', header = T, row.names = 1)
roc_pvalues <- read.csv('../results/tables/roc_pvalue_summary.csv', header = T, row.names = 1)

# Predictive of both initial and follow up and in lesion model
pvalue_common_imp_vars <- read.csv('../results/tables/pvalue_common_imp_vars.csv', header = T, row.names = 1)

@@

Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.  Previous research suggests that the bacterial microbiome can be used as a biomarker for colorectal cancer (CRC).  These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.  In this study, we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for lesion (adenoma or carcinoma) after surgical treatment.  This model was tested on a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment. The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.  : For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```).  In the test set, the prediction of lesion for the initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).  Our model predicted that ```r format((followup_cm["pred_no", "ref_no"]/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% of the 67-person cohort had normal colons and ```r format(((followup_cm["pred_yes", "ref_no"] + sum(followup_cm$ref_yes))/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% had a lesion.  Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

## Discussion
In this study we show that the overall community structure as measured by different alpha diversity metrics, shows very little change between controls and those with either adenoma or carcinoma [Table S1] in our larger training set.  With respect to the follow up sample group there was very little difference in magnitude of change in the thetayc distance metric between those with adenoma or carcinoma[Figure 1a].  FIT had a large change in the initial and follow up samples in the carcinoma group versus the adenoma [Figure 1b].  An NMDS showed that there was very little change between initial and follow up for the adenoma group but there was a change in the carcinoma group [Figure 2].  This information showed that the removal and treatment of carcinoma for these cursory metrics had the largest response.  Yet these metrics alone would not be able to identify differences in adenomas.

### 2016-12-05 submission/manuscript_outline_20161024.tex
updated manuscript w/ newer abstract

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests that the bacterial microbiome
can be used as a biomarker for colorectal cancer (CRC). These reports
have mostly focused on investigating how the bacterial microbiome is
used at a single point in time to predict disease. In this study, we
assessed whether a model built with bacterial microbiome data could
accurately predict adenoma or carcinoma and adjust positive probability
for lesion (adenoma or carcinoma) after surgical treatment. This model
was tested on a \texttt{66} person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model chosen had a 10-fold cross validated AUC
of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
up samples our Random Forest model significantly decreased the positive
probability of lesion compared to the initial samples for both adenoma
(P-value = \texttt{8.48e-07}) and carcinoma (P-value =
\texttt{5.96e-08}). The top 5 most important operational taxonomic units
(OTUs) for prediction of lesion classified to
\texttt{Lachnospiraceae\ (Otu000013)},
\texttt{Escherichia/Shigella\ (Otu000018)},
\texttt{Ruminococcaceae\ (Otu000020)},
\texttt{Ruminococcus\ (Otu000017)}, and
\texttt{Porphyromonas\ (Otu000153)}. In the test set, the prediction of
lesion for the initial samples had a sensitivity of \texttt{98.5}\%
(\texttt{65/66}) while follow up it was \texttt{100}\%
(\texttt{1\ /\ 1}). Overall, our model accurately predicted those with
an adenoma or carcinoma and following treatment it also decreased the
positive probability of having a lesion.

@@

In this study we show that the overall community structure shows very
little change between controls and those with either adenoma or
carcinoma {[}Table S1{]} in our larger training set. With respect to the
follow up sample group there was very little difference in thetayc
distance metric between those with adenoma or carcinoma{[}Figure 1a{]}.
FIT had a large change in the initial and follow up samples in the
carcinoma group versus the adenoma {[}Figure 1b{]}. When investigating
the bacterial community, specifically between these individuals, there
was very little change in overall community in the adenoma group but
there was a change in the carcinoma group {[}Figure 2{]}. This
information showed that the removal and treatment of carcinoma for these
cursory metrics had the largest response. Yet these metrics alone would
not be able to identify differences in adenomas.

>>>>

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests that the bacterial microbiome
can be used as a biomarker for colorectal cancer (CRC). These reports
have mostly focused on investigating how the bacterial microbiome is
used at a single point in time to predict disease. In this study, we
assessed whether a model built with bacterial microbiome data could
accurately predict adenoma or carcinoma and adjust positive probability
for lesion (adenoma or carcinoma) after surgical treatment. This model
was tested on a \texttt{66} person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model chosen had a 10-fold cross validated AUC
of \texttt{0.819} and a test set AUC of \texttt{0.815}. : For the follow
up samples our Random Forest model significantly decreased the positive
probability of lesion compared to the initial samples for both adenoma
(P-value = \texttt{8.48e-07}) and carcinoma (P-value =
\texttt{5.96e-08}). In the test set, the prediction of lesion for the
initial samples had a sensitivity of \texttt{98.5}\% (\texttt{65/66})
while follow up it was \texttt{100}\% (\texttt{1\ /\ 1}). Our model
predicted that \texttt{39.4}\% of the 67-person cohort had normal colons
and \texttt{60.6}\% had a lesion. Our model suggests that treatment does
significantly reduce the probability of having a colonic lesion. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

In this study we show that the overall community structure as measured
by different alpha diversity metrics, shows very little change between
controls and those with either adenoma or carcinoma {[}Table S1{]} in
our larger training set. With respect to the follow up sample group
there was very little difference in magnitude of change in the thetayc
distance metric between those with adenoma or carcinoma{[}Figure 1a{]}.
FIT had a large change in the initial and follow up samples in the
carcinoma group versus the adenoma {[}Figure 1b{]}. An NMDS showed that
there was very little change between initial and follow up for the
adenoma group but there was a change in the carcinoma group {[}Figure
2{]}. This information showed that the removal and treatment of
carcinoma for these cursory metrics had the largest response. Yet these
metrics alone would not be able to identify differences in adenomas.
